MAESTROeX: A Massively Parallel Low Mach Number Astrophysical Solver

### 2019-06-07 MAESTROeX_I/algorithm.tex
sec 4 readthrough
typos, spell check

In this paper we explore a new option to retain an irregularly-spaced base state to eliminate mapping errors from the ``fill'' module.
For the lateral average, as before we first sum the values in all the cells associated with each radius and divide
by the number of contributing cells to obtain the arithmetic average.  However we do not map this onto a regularly spaced
base state and retain the use of this irregularly-spaced base state.
The advantage with this approach is that the fill module does not require any interpolation.
A potential benefit to eliminating the mapping error is to consider a spherical star in hydrostatic equlibrium at rest.
In the absense of reactions, the star should remain at rest.
The buoyancy forcing term in the momentum equation contains $\rho-\rho_0$.  With the original scheme, interpolation errors 
in computing $\rho_0$ by averaging would cause artificial acceleration in the velocity field due to the interpolation error 
from the Cartesian grid to and from the radial base state.  By retaining the radial base state as an irregularly spaced 
array, the effects due to interpolation error are nearly eliminated, but not completely eliminated since there are still 
machine precision effects resulting from averaging a large number of numerical values.
We note that $\Delta r$ decreases as the base state moves further from the center of the star, 
which results in far more total cells in the irregularly-spaced array than the previous uniformly-spaced array.

@@

There are a few key numerical modules we use in each time step.
\begin{itemize}
\item {\bf Average}$[\phi]\rightarrow[\overline\phi]$ computes the lateral average of a quantity over a layer at constant radius $r$, as described above in Section \ref{Sec:Spatial}.
\item {\bf Enforce HSE}$[\rho_0]\rightarrow[p_0]$ computes the base state pressure, $p_0$, from a base state density, $\rho_0$ by integrating the hydrostatic equilibrium condition in one dimension. 
This follows equation (A10) in Paper V, noting that for the irregularly spaced base state case, $\Delta r$ is not constant, where $\Delta r_{j+1/2} = r_{j+1} - r_j$ for cell-edge $j+1/2$.  
The base state pressure remains equal to a constant value at the location of a prescribed cutoff density outward for the entire simulation.
\item {\bf React State}$[(\rho X_k)^{\rm in},(\rho h)^{\rm in},p_0]\rightarrow[(\rho X_k)^{\rm out},(\rho h)^{\rm out},(\rho\dot\omega),(\rho\Hnuc)]$ integrates the species and enthalpy due to reactions over $\Delta t/2$ by solving
\begin{equation}
\frac{dX_k}{dt} = \dot\omega_k(\rho,X_k,T); \qquad
\frac{dT}{dt} = \frac{1}{c_p}\left(-\sum_k\xi_k\dot\omega_k + H_{\rm nuc})\right)
\end{equation}
The inputs are the species, enthalpy, and base state pressure, and the outpus are the species, enthalpy, reaction rates, and nuclear energy generation rate.
See Paper III for details.
\end{itemize}
Each time step is constrained by the standard advective CFL condition,
\begin{equation}
\Delta t = \sigma^{\rm CFL} \min_i(\Delta x / U_i),
\end{equation}
where for our simulations we typically use $\sigma^{\rm CFL}\sim 0.7$ and the minimum is taken over all spatial directions over all cells.
There are additional constraints on the timestep that are typically much less restrictive than the advective CFL including
the acceleration due to the buoyancy force (sometimes in effect when the velocity is approximately zero at the start of some simulations) 
and the local magnitude of the divergence constraint (to prevent too much mass evacuation from a cell in a time step); see Section 3.4 in Paper III for details.

@@

Our approach to adaptive mesh refinement is algorithmically the same as the treatment described in Section 5 of Paper V; we refer the reader there for details.
We note that for spherical problems, AMR is only available for the case of a uniformly-spaced base state.

>>>>

In this paper we explore a new option to retain an irregularly-spaced base state to eliminate mapping errors from the ``fill'' module.
For the lateral average, as before we first sum the values in all the cells associated with each radius and divide
by the number of contributing cells to obtain the arithmetic average.  However we do not map this onto a regularly spaced
base state and retain the use of this irregularly-spaced base state.
The advantage with this approach is that the fill module does not require any interpolation.
A potential benefit to eliminating the mapping error is to consider a spherical star in hydrostatic equilibrium at rest.
In the absence of reactions, the star should remain at rest.
The buoyancy forcing term in the momentum equation contains $\rho-\rho_0$.  With the original scheme, interpolation errors 
in computing $\rho_0$ by averaging would cause artificial acceleration in the velocity field due to the interpolation error 
from the Cartesian grid to and from the radial base state.  By retaining the radial base state as an irregularly spaced 
array, the effects due to interpolation error are nearly eliminated, but not completely eliminated since there are still 
machine precision effects resulting from averaging a large number of numerical values.
We note that $\Delta r$ decreases as the base state moves further from the center of the star, 
which results in far more total cells in the irregularly-spaced array than the previous uniformly-spaced array.

@@

There are a few key numerical modules we use in each time step.
\begin{itemize}
\item {\bf Average}$[\phi]\rightarrow[\overline\phi]$ computes the lateral average of a quantity over a layer at constant radius $r$, as described above in Section \ref{Sec:Spatial}.
\item {\bf Enforce HSE}$[\rho_0]\rightarrow[p_0]$ computes the base state pressure, $p_0$, from a base state density, $\rho_0$ by integrating the hydrostatic equilibrium condition in one dimension. 
This follows equation (A10) in Paper V, noting that for the irregularly spaced base state case, $\Delta r$ is not constant, where $\Delta r_{j+1/2} = r_{j+1} - r_j$ for cell-edge $j+1/2$.  
The base state pressure remains equal to a constant value at the location of a prescribed cutoff density outward for the entire simulation.
\item {\bf React State}$[(\rho X_k)^{\rm in},(\rho h)^{\rm in},p_0]\rightarrow[(\rho X_k)^{\rm out},(\rho h)^{\rm out},(\rho\dot\omega),(\rho\Hnuc)]$ integrates the species and enthalpy due to reactions over $\Delta t/2$ by solving
\begin{equation}
\frac{dX_k}{dt} = \dot\omega_k(\rho,X_k,T); \qquad
\frac{dT}{dt} = \frac{1}{c_p}\left(-\sum_k\xi_k\dot\omega_k + H_{\rm nuc})\right)
\end{equation}
The inputs are the species, enthalpy, and base state pressure, and the outputs are the species, enthalpy, reaction rates, and nuclear energy generation rate.
See Paper III for details.
\end{itemize}
Each time step is constrained by the standard advective CFL condition,
\begin{equation}
\Delta t = \sigma^{\rm CFL} \min_i(\Delta x / U_i),
\end{equation}
where for our simulations we typically use $\sigma^{\rm CFL}\sim 0.7$ and the minimum is taken over all spatial directions over all cells.
There are additional constraints on the time step that are typically much less restrictive than the advective CFL including
the acceleration due to the buoyancy force (sometimes in effect when the velocity is approximately zero at the start of some simulations) 
and the local magnitude of the divergence constraint (to prevent too much mass evacuation from a cell in a time step); see Section 3.4 in Paper III for details.

@@

Our approach to adaptive mesh refinement is algorithmically the same as the treatment described in Section 5 of Paper V; we refer the reader there for details.
MAESTROeX supports refinement ratios of 2 between levels.
We note that for spherical problems, AMR is only available for the case of a uniformly-spaced base state.

### 2019-06-07 MAESTROeX_I/appendix.tex
typos, spell check

\appendix
\section{Projection Details}\label{Sec:Projection}
To enforce the divergence constraint in {\bf Step 3, 7}, and {\bf 11}, we use a projection method analgous to the methods originally developed for incompressible flow \citep{almgren1998conservative,bell1989second}.
The basic idea is to decompose the velocity field into a divergence-free component and a curl-free (gradient of a scalar field) component by solving a variable-coefficient Poisson equation for the scalar field.
The details for the MAC projection in {\bf Step 3} and {\bf Step 7} are given in Appendix B of Paper III.  The details of the nodal projection in {\bf Step 11} are given in Section 3.2 of Paper III.
We note that in the nodal projection, the gradient of the scalar field is used to update the perturbational pressure, $\pi$.

>>>>

\appendix
\section{Projection Details}\label{Sec:Projection}
To enforce the divergence constraint in {\bf Step 3, 7}, and {\bf 11}, we use a projection method analogous to the methods originally developed for incompressible flow \citep{almgren1998conservative,bell1989second}.
The basic idea is to decompose the velocity field into a divergence-free component and a curl-free (gradient of a scalar field) component by solving a variable-coefficient Poisson equation for the scalar field.
The details for the MAC projection in {\bf Step 3} and {\bf Step 7} are given in Appendix B of Paper III.  The details of the nodal projection in {\bf Step 11} are given in Section 3.2 of Paper III.
We note that in the nodal projection, the gradient of the scalar field is used to update the perturbational pressure, $\pi$.

### 2019-06-07 MAESTROeX_I/conclusions.tex
typos, spell check

\section{Conclusions and Future Work}\label{sec:conclusions}
We have developed a new temporal integrator and spatial mapping options into our low Mach number solver, MAESTROeX.
The new AMReX-enabled code scales well on large fractions of supercomputers with multicore architectures.
Future software enhancements will include GPU implementation.
In particular, the AMReX-based companion code, the compressible CASTRO code \citep{CASTRO}, has recently ported hydrodynamics and reactions to GPUs \citep{CASTRO_GPU}.
We plan to leverage the newly implemented mechanisms for offloading compute kernels to GPUs inside of the AMReX software library itself.
Our future scientific investigations include convection in massive rotating stars, \citep{heger2000presupernova}, the convective Urca process in white dwarfs \citep{willcox2016type}, solar physics \citep{wood2018self}, and magnetohydrodynamics \citep{wood2015three,wood2011sun}.
Our future algorithmic enhancements include more accurate and higher-order multiphysics coupling strategies byased on spectral deferred corrections \citep{dutt2000spectral,bourlioux2003high}.
This framework has been successfully used in terrestrial combustion \citep{pazner2016high,nonaka2018conservative}

>>>>

\section{Conclusions and Future Work}\label{sec:conclusions}
We have developed a new temporal integrator and spatial mapping options into our low Mach number solver, MAESTROeX.
The new AMReX-enabled code scales well on large fractions of supercomputers with multicore architectures.
Future software enhancements will include GPU implementation.
In particular, the AMReX-based companion code, the compressible CASTRO code \citep{CASTRO}, has recently ported hydrodynamics and reactions to GPUs \citep{CASTRO_GPU}.
We plan to leverage the newly implemented mechanisms for offloading compute kernels to GPUs inside of the AMReX software library itself.
Our future scientific investigations include convection in massive rotating stars, \citep{heger2000presupernova}, the convective Urca process in white dwarfs \citep{willcox2016type}, solar physics \citep{wood2018self}, and magnetohydrodynamics \citep{wood2015three,wood2011sun}.
Our future algorithmic enhancements include more accurate and higher-order multiphysics coupling strategies based on spectral deferred corrections \citep{dutt2000spectral,bourlioux2003high}.
This framework has been successfully used in terrestrial combustion \citep{pazner2016high,nonaka2018conservative}

### 2019-06-07 MAESTROeX_I/equations.tex
typos, spell check

Mathematically, equations (\ref{eq:momentum})-(\ref{eq:enthalpy}) must still be closed by the EOS.  
This is done by taking the Langrangian derivative of the EOS for pressure as a function of the thermodynamic variables, 
substituting in the equations of motion for mass and energy,
and requiring that the pressure is a prescribed function of altitude and time based on the hydrostatic equilibrium condition.
See Papers I and II for details of this derivation.
The resulting equation is a divergence constraint on the velocity field,
\begin{equation}
\nabla\cdot(\beta_0\Ub) = \beta_0\left(S - \frac{1}{\gammaonebar p_0}\frac{\partial p_0}{\partial t}\right).\label{eq:U divergence}
\end{equation}
The expansion term, $S$, incorporates local compressibility effects due to compositional changes and heat release from reactions,
\begin{equation}
S = -\sigma\sum_k\xi_k\omegadot_k + \frac{1}{\rho p_\rho}\sum_k p_{X_k}\omegadot_k + \sigma\Hnuc,\label{eq:S}
\end{equation}
where 
$p_{X_k} \equiv \left. \partial p / \partial X_k \right|_{\rho,T,X_{j,j\ne k}}$, 
$\xi_k \equiv \left. \partial h /\partial X_k \right |_{p,T,X_{j,j\ne k}}$,
$p_\rho \equiv \left.\partial p/\partial \rho \right |_{T, X_k}$, and 
$\sigma \equiv p_T/(\rho c_p p_\rho)$, with $p_T \equiv \left. \partial p / \partial
T \right|_{\rho, X_k}$ and $c_p \equiv \left.  \partial h / \partial T
\right|_{p,X_k}$ is the specific heat at constant pressure.

>>>>

Mathematically, equations (\ref{eq:momentum})-(\ref{eq:enthalpy}) must still be closed by the EOS.  
This is done by taking the Lagrangian derivative of the EOS for pressure as a function of the thermodynamic variables, 
substituting in the equations of motion for mass and energy,
and requiring that the pressure is a prescribed function of altitude and time based on the hydrostatic equilibrium condition.
See Papers I and II for details of this derivation.
The resulting equation is a divergence constraint on the velocity field,
\begin{equation}
\nabla\cdot(\beta_0\Ub) = \beta_0\left(S - \frac{1}{\gammaonebar p_0}\frac{\partial p_0}{\partial t}\right).\label{eq:U divergence}
\end{equation}
The expansion term, $S$, incorporates local compressibility effects due to compositional changes and heat release from reactions,
\begin{equation}
S = -\sigma\sum_k\xi_k\omegadot_k + \frac{1}{\rho p_\rho}\sum_k p_{X_k}\omegadot_k + \sigma\Hnuc,\label{eq:S}
\end{equation}
where 
$p_{X_k} \equiv \left. \partial p / \partial X_k \right|_{\rho,T,X_{j,j\ne k}}$, 
$\xi_k \equiv \left. \partial h /\partial X_k \right |_{p,T,X_{j,j\ne k}}$,
$p_\rho \equiv \left.\partial p/\partial \rho \right |_{T, X_k}$, and 
$\sigma \equiv p_T/(\rho c_p p_\rho)$, with $p_T \equiv \left. \partial p / \partial
T \right|_{\rho, X_k}$ and $c_p \equiv \left.  \partial h / \partial T
\right|_{p,X_k}$ is the specific heat at constant pressure.

### 2019-06-07 MAESTROeX_I/intro.tex
typos, spell check

Low Mach number models have been developed for a variety of contexts
including combustion \citep{day2000numerical}, terrestrial atmospheric
modeling \citep{durran:1989,oneill:2014,duarte2015low}, and elastic
solids \citep{abbate2017all}.  For astrophysical applications, a
number of approaches to modeling low Mach number flows have been
developed in recent years.  One of the more similar approaches to ours is
\cite{Lin:2006}, however this approach is only first-order accurate and does
not account for atmospheric expansion.  There are also semi-implicit all-Mach
number solvers, where the Euler equations are split into an acoustic
part and an advective part
\citep{Kwatra2009,Degond2009,Cordier2012,Haack2012,Happenhofer2013,Chalons2016,Padioleau2019}.
The fast acoustic waves are then solved using implicit time
integration, while the slow material waves are solved explicitly.
Another approach is to use preconditioned all-Mach number solvers
\citep{Miczek2014,Barsukow2016}, where the numerical flux is
multiplied by a preconditioning matrix.  This reduces the stiffness of
the system at low Mach numbers, while retaining the correct scaling
behavior. In the reduced speed of sound technique (RSST) and related
methods, the speed of sound is artificially reduced by including a
suitable scaling factor in the continuity equation, reducing the
restriction on the size of the timestep
\citep{Rempel2005,Hotta2012,Wang2015,Takeyama2017,Iijima2018}.
Finally, there are fully implicit time integration codes for the
compressible Euler equations
\citep{Viallet2011,kifonidis:2012,Viallet2015,Goffrey2016}.  The MUSIC code
uses fully implicit time integration for the compressible Euler
equations, which therefore allows for arbitrarily large timesteps.

Previously, we developed the low Mach number astrophysical solver, MAESTRO.
MAESTRO is a block-structured, Cartesian grid finite-volume, adaptive mesh refinement (AMR) 
code that has been successfully used for many years for a number of applications, detailed below.
Unlike several of the references above, MAESTRO is not an all-Mach solver, but is suitable for 
flows where the Mach number is small ($\sim 0.1$ or smaller).  
Furthermore, the low Mach number model in MAESTRO is specifically designed for, but not limited 
to, astrophysical settings with significant atmospheric stratification.
This includes full spherical stars, as well as planar simulations of dynamics within localized 
regions of a star.
The numerical methodology relies on an explicit Godunov approach for advection, a stiff ODE 
solver for reactions (VODE) \citep{vode}, and multigrid-based linear solvers for the 
pressure-projection steps.  This, the time step is limited by an advective CFL constraint based on the 
fluid velocity, not the sound speed.
Central to the algorithm are time-varying, one-dimensional stratified background (or base) state 
density and pressure fields that are held in hydrostatic equilibrium.
The base state density couples to the full state solution through buyoancy terms in the momentum equation,
and the base state pressure couples to the full state solution by constraining the evolution of the
thermodynamic variables to match this pressure.
The time-advancement strategy uses a predictor-corrector strategy for evolving the base state,
Strang splitting to integrate the thermodynamic variables, and a 
second-order projection method to integrate the velocity equations subject to a divergence constraint.
The original MAESTRO code was developed in the pure-Fortran 90 FBoxLib software framework, whereas
MAESTROeX is developed in the C++/F90 AMReX framework \citep{AMReX,AMReX_JOSS}.

>>>>

Low Mach number models have been developed for a variety of contexts
including combustion \citep{day2000numerical}, terrestrial atmospheric
modeling \citep{durran:1989,oneill:2014,duarte2015low}, and elastic
solids \citep{abbate2017all}.  For astrophysical applications, a
number of approaches to modeling low Mach number flows have been
developed in recent years.  One of the more similar approaches to ours is
\cite{Lin:2006}, however this approach is only first-order accurate and does
not account for atmospheric expansion.  There are also semi-implicit all-Mach
number solvers, where the Euler equations are split into an acoustic
part and an advective part
\citep{Kwatra2009,Degond2009,Cordier2012,Haack2012,Happenhofer2013,Chalons2016,Padioleau2019}.
The fast acoustic waves are then solved using implicit time
integration, while the slow material waves are solved explicitly.
Another approach is to use preconditioned all-Mach number solvers
\citep{Miczek2014,Barsukow2016}, where the numerical flux is
multiplied by a preconditioning matrix.  This reduces the stiffness of
the system at low Mach numbers, while retaining the correct scaling
behavior. In the reduced speed of sound technique (RSST) and related
methods, the speed of sound is artificially reduced by including a
suitable scaling factor in the continuity equation, reducing the
restriction on the size of the time step
\citep{Rempel2005,Hotta2012,Wang2015,Takeyama2017,Iijima2018}.
Finally, there are fully implicit time integration codes for the
compressible Euler equations
\citep{Viallet2011,kifonidis:2012,Viallet2015,Goffrey2016}.  The MUSIC code
uses fully implicit time integration for the compressible Euler
equations, which therefore allows for arbitrarily large time steps.

Previously, we developed the low Mach number astrophysical solver, MAESTRO.
MAESTRO is a block-structured, Cartesian grid finite-volume, adaptive mesh refinement (AMR) 
code that has been successfully used for many years for a number of applications, detailed below.
Unlike several of the references above, MAESTRO is not an all-Mach solver, but is suitable for 
flows where the Mach number is small ($\sim 0.1$ or smaller).  
Furthermore, the low Mach number model in MAESTRO is specifically designed for, but not limited 
to, astrophysical settings with significant atmospheric stratification.
This includes full spherical stars, as well as planar simulations of dynamics within localized 
regions of a star.
The numerical methodology relies on an explicit Godunov approach for advection, a stiff ODE 
solver for reactions (VODE) \citep{vode}, and multigrid-based linear solvers for the 
pressure-projection steps.  This, the time step is limited by an advective CFL constraint based on the 
fluid velocity, not the sound speed.
Central to the algorithm are time-varying, one-dimensional stratified background (or base) state 
density and pressure fields that are held in hydrostatic equilibrium.
The base state density couples to the full state solution through buoyancy terms in the momentum equation,
and the base state pressure couples to the full state solution by constraining the evolution of the
thermodynamic variables to match this pressure.
The time-advancement strategy uses a predictor-corrector strategy for evolving the base state,
Strang splitting to integrate the thermodynamic variables, and a 
second-order projection method to integrate the velocity equations subject to a divergence constraint.
The original MAESTRO code was developed in the pure-Fortran 90 FBoxLib software framework, whereas
MAESTROeX is developed in the C++/F90 AMReX framework \citep{AMReX,AMReX_JOSS}.
